---
title: "Understanding Sequencing Raw Data"
output: html_document
---

# Understanding Sequencing Raw Data

## Class Environment

### Getting into AWS Instance

```{bash, eval = FALSE}
## Example
ssh -i berlin.pem ubuntu@<PUBLIC IP ADDRESS> (e.g.34.219.254.245)

## Actual Command
ssh i berlin.pem ubuntu@34.213.180.241
```

<!--
### Rstudio

```{bash, eval = FALSE, echo = FALSE}
## maybe take away the --rm so they can save the container for later
## or whatever port
docker run --rm -it -e PASSWORD=train -p 9000:8787 kdgosik/scellbern2019

sudo docker run --rm -it -e PASSWORD=train -v $PWD/Share:/data -v $PWD/userfolders/Dana:/mydir kdgosik/scellbern2019
```
-->

## Shell and Unix commands
<iframe src="https://docs.google.com/presentation/d/e/2PACX-1vSavPAPY3WXe2EiqaRXJmAajFTeGcRbD1qJq4Sp4YlcF1_m7I1X89D1uC-8jDDGinqFLNO2oJNALbNx/embed?start=false&loop=false&delayms=3000" frameborder="0" width="760" height="569" allowfullscreen="true" mozallowfullscreen="true" webkitallowfullscreen="true"></iframe>


### Common Linux Commands

#### Lab 1a

  - check the your present directory
```{r, eval = FALSE}
pwd
```

  - make a directory called data
```{r, eval = FALSE}
mkdir data
```

  - change into data directory
```{r, eval = FALSE, echo = FALSE}
cd data
```

  - check manual page of wget command
```{bash, eval = FALSE, echo = FALSE}
man wget
```

  - redirect wget maunual page output into a file called wget.txt
```{bash, eval = FALSE, echo = FALSE}
man wget > wget.txt
```

  - return the lines that contain output in the wget.txt file
```{bash, eval = FALSE, echo = FALSE}
cat wget.txt | grep output
```

```{bash, eval = FALSE, echo = FALSE}
grep -i output wget.txt
```

  - Compress wget.txt file
```{bash, eval = FALSE, echo = FALSE}
gzip wget.txt
```

  - View Compressed file
```{bash, eval = FALSE, echo = FALSE}
cat wget.txt.qz
```
  

```{bash, eval = FALSE, echo = FALSE}
zcat wget.txt.qz
```


```{bash, eval = FALSE, echo = FALSE}
zcat wget.txt.qz | less
```





#### Docker Commands

```{bash, eval = FALSE, echo = FALSE}
## maybe take away the --rm so they can save the container for later
## or whatever port
docker run --rm -it -e PASSWORD=train -p 9000:8787 kdgosik/scellbern2019 bash

sudo docker run --rm -it -e PASSWORD=train -v $PWD/Share:/data kdgosik/scellbern2019 bash

sudo docker run --rm -it -e PASSWORD=train -v $PWD/Share:/data -v $PWD/userfolders/Dana:/mydir kdgosik/scellbern2019 bash
```

```{bash, eval = FALSE}
## maybe take away the --rm so they can save the container for later

sudo docker run --rm -it -v $PWD/Share:/Share kdgosik/scellbern2019 bash
```

**Explaination of commands**
  - docker: command to run docker
  - run: asking docker to run a container
  - --rm: flag to remove the container when you exit from it
      - nothing will be saved from your session to access again later
      - this flag can be removed to keep container
  - -it: flag to run the container interactively
    - this will keep all session output displaying on the terminal
    - to stop container go to terminal and press Crtl+c
  - kdgosik/scellbern2019: the image to run.  It will be the image into a container if not already built on your computer
    - [image link](https://hub.docker.com/r/kdgosik/scellbern2019)


## File formats

  - bcl
  - fastq
  - bam
  - mtx, tsv
  - hdf5 (.h5, .h5ad)

## Public data repositories

### Cellranger/10x

#### Lab 1b

10x PBMC data are hosted in https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz


  - change directory into the data directory
  - get 10x PBMC data
  - unzip data
  - explore directory
  - explore files


```{bash, echo = FALSE, eval = FALSE}
mkdir data
wget https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz -O data/pbmc3k_filtered_gene_bc_matrices.tar.gz
cd data; tar -xzf pbmc3k_filtered_gene_bc_matrices.tar.gz
cd ..
```




### GEO

 - https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE81905


#### Lab 1c

**Get GEO Data**
  - ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81905/matrix/GSE81905-GPL19057_series_matrix.txt.gz
  - ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81905/matrix/GSE81905-GPL17021_series_matrix.txt.gz


 - make a directory for the files or use data directory
 - go into that directory
 - get files and place them in the directory
 - View files (try keeping in compressed format and view that way)
 
 
### Single Cell Portal

  - https://portals.broadinstitute.org/single_cell

  - Study: Salk Institute - Single-cell Methylome Sequencing Identifies Distinct Neuronal Populations in Mouse Frontal Cortex

#### Lab 1d

  - Get R2 fastq file from the Salk Institute study 
  - Look at files
  
  
#### Lab 1e

  - Get Docker on your local computer for you to have 
  - Explore Single Cell Portal
  - Explore GEO
